Robust Digital Envelope Estimation Via Geometric Properties of an Arbitrary Real Signal

### 2020-11-23 Papers/01_Pulses_Envelope/PulsesEnvelope.tex
moved files to make rooms for new developments
revising "pulses envelope"

\documentclass{article}
\usepackage{lmodern} % Latin Modern Font
\usepackage[T1]{fontenc}
\usepackage{authblk}
\usepackage{amsfonts} % \mathbb
\usepackage[margin=0.5in]{geometry}
\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage[backend=biber, style=apa]{biblatex}
\usepackage{amsmath}
\usepackage{hyperref}
\usepackage{hypcap}
\DeclareMathOperator{\sign}{sign}
\bibliography{bibli}
\hypersetup{
  colorlinks=true, 
  allcolors=black,
  pdfauthor={Carlos Henrique Tarjano Santos},
  pdftitle={Robust Digital Envelope Estimation Via Geometric Properties of an Arbitrary Real Signal}
}
\title{Robust Digital Envelope Estimation Via Geometric Properties of an Arbitrary Real Signal}
\author[1]{Carlos Henrique Tarjano Santos}
\author[2]{Valdecy Pereira}
\affil[1]{(corresponding author, carlostarjano@id.uff.br) Department of Production Engineering, Universidade Federal Fluminense, Rua Passo da Pátria, 156, Campus Praia Vermelha, Bloco D - sala 309, São Domingos, Niterói, RJ, Brasil, CEP: 24.210-240}
\affil[2]{Department of Production Engineering, Universidade Federal Fluminense, Rua Passo da Pátria, 156, Campus Praia Vermelha, Bloco D - sala 309, São Domingos, Niterói, RJ, Brasil, CEP: 24.210-240}

@@

\section{Introduction}
Envelope detection is ubiquitous in both analogue and digital signal processing \parencite{2011CaetanoImproved}. Nevertheless, the literature in this area is very fragmented \parencite{2017LyonsDigital}. Besides, most envelope detection techniques are designed to account for very specific settings, like pure sinusoids with moderate noise content, a limitation that excludes most physical signals, as is the case of recorded sound, for example; that limitation arises in part due to the lack of a strict mathematical definition of a temporal envelope \parencite{2013Mengempirical}.

In many natural signals, however, the temporal amplitude envelope of a signal plays a prominent role in the characteristics exhibited: according to \textcite{2017QiRelative}, for example, the envelope is at least as important as the fine structure of a sound wave in the context of the intelligibility of mandarin tones. This is also the case for English \parencite{1995ShannonSpeech}, where even envelopes modulating mostly noise were still capable of conveying meaning.

The envelope helps to impart emotion and identity to the human voice \parencite{2018ZhuContributions}, and envelope preserving characteristics in concert halls are associated with their pleasantness \parencite{2011LokkiEngaging}.

@@

Addressing the different units in the horizontal and vertical axes, one can transform the DSP problem of envelope detection into the geometric problem of defining the shape of a set of points in $ \mathbb{R}^2 $. For that purpose, \textcite{1983Edelsbrunnershape} introduced the concept of alpha-shapes, a mathematically well-defined extension to the convex hull of a finite set of points, closely related to the Delaunay triangulation and Voronoi diagrams of those points.

This approach is used in areas such as the detection of features in images \parencite{2016VarytimidisAlpha}, reconstruction of surfaces from a cloud of points \parencite{2015WuAutomated} and Spectroscopy \parencite{2019XuModeling}, with the last work, which involves the estimation and removal of the Blaze function (a kind of envelope) of an echelle spectrograph, being particularly illustrative of the potential of cross-fertilization.

Steps in the direction of translating geometric algorithms to the context of envelope detection were also made by \textcite{2015YangSkeleton} via an algorithm based on the construction of a skeleton underlying the wave, and also via the direct translation of computer vision methods to the task \parencite{2015YangRepresenting}.

Following this path, we present a general approach to envelope detection, exploiting the intrinsic characteristics of a generic, spectrally complex wave, in order to avoid the need for manual intervention or parameter tuning.

@@

\section{Methodology}

In this work it is assumed, without loss of generality, that one is interested in retrieving the mathematically ill-defined \parencite{2013Mengempirical} temporal envelope of an arbitrary physical sound wave, of which only a digital representation is available.

The best representation one can hope to get of the original wave is, then, the recording itself. It is sensible, thus, to characterize the envelope of the wave in terms of a subset of the exact samples that comprise its digital representation.

@@

  \begin{figure}[ht!]
    \centering
      \includegraphics[width=0.8\linewidth]{01Pulses.pdf}
    \caption{Example of a discrete wave. The black stems represent the individual samples of the signal. The grey areas encompass the simplified convex hulls of the pulses. The dashed lines mark the frontiers between pseudo-cycles.}
    \label{fig:Pulses}
  \end{figure}

The most mathematically sound definition of an envelope involves the representation of its underlying wave as an analytic signal, introduced by Gabor in 1946 \parencite{2007HahnHistory}. In his work, \textcite{1946GaborTheory} applies the then relatively new mathematical machinery of the quantum mechanics to unify the time and frequency representations of a wave, showing how the Hilbert transform could be applied to a real signal in order to obtain an equivalent complex signal, later known as the analytic signal.

Figure \ref{fig:AnalyticSignal} shows the envelope obtained via the analytic signal, for a pure sinusoid with a local frequency of 20 cycles, modulated by a polynomial of the third degree, illustrating the difference in the shape of the envelope in the presence and absence of gaussian white noise with standard deviation of $ 1/10 $ of the wave's maximum amplitude.

\begin{figure}[ht!]
  \centering
    \includegraphics[width=0.8\linewidth]{02AnalyticSignal.pdf}
  \caption{Envelope of a pure sinusoid modulated by a polynomial of degree 3, as obtained by the Hilbert Transform. In the first half, the sinusoid is free of noise, while in the second half white gaussian noise with a standard deviation of $ 1/10 $ of the maximum amplitude of the wave was added to the base signal.}
  \label{fig:AnalyticSignal}
\end{figure}

It is readily noticeable from figure \ref{fig:AnalyticSignal} that, as soon as noise is introduced in the original signal, it is reflected in the envelope, breaking our assumption of smoothness. Other than that, this definition yields a unique, positive envelope.

>>>>

\documentclass{article}
\usepackage{lmodern} % Latin Modern Font
\usepackage[T1]{fontenc}
\usepackage{authblk}
\usepackage{amsfonts} % \mathbb
\usepackage[margin=0.5in]{geometry}
\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage[backend=biber]{biblatex}
\usepackage{amsmath}
% \usepackage{hyperref}
% \usepackage{hypcap}
\DeclareMathOperator{\sign}{sign}
\bibliography{bibli}
% \hypersetup{
%   colorlinks=true, 
%   allcolors=black,
%   pdfauthor={Carlos Henrique Tarjano Santos},
%   pdftitle={Robust Digital Envelope Estimation Via Geometric Properties of an Arbitrary Real Signal}
% }
\title{Robust Digital Envelope Estimation Via Geometric Properties of an Arbitrary Real Signal}
\author[1]{Carlos Henrique Tarjano Santos}
\author[2]{Valdecy Pereira}
\affil[1]{(corresponding author, carlostarjano@id.uff.br) Department of Production Engineering, Universidade Federal Fluminense, Rua Passo da Pátria, 156, Campus Praia Vermelha, Bloco D - sala 309, São Domingos, Niterói, RJ, Brasil, CEP: 24.210-240}
\affil[2]{Department of Production Engineering, Universidade Federal Fluminense, Rua Passo da Pátria, 156, Campus Praia Vermelha, Bloco D - sala 309, São Domingos, Niterói, RJ, Brasil, CEP: 24.210-240}

@@

\section{Introduction}
Envelope detection, also known as demodulation, is ubiquitous in both analogue and digital signal processing \parencite{2011CaetanoImproved}. Nevertheless, the literature in this area is very fragmented \parencite{2017LyonsDigital}. Besides, most envelope detection techniques are designed to account for very specific settings, like pure sinusoids with moderate noise content, a limitation that excludes most physical signals, as is the case of recorded sound, for example; that limitation arises in part due to the lack of a strict mathematical definition of a temporal envelope \parencite{2013Mengempirical}.

In many contexts, however, the temporal amplitude envelope of a signal plays a prominent role: according to \textcite{2017QiRelative}, for example, the envelope is at least as important as the fine structure of a sound wave in the context of the intelligibility of Mandarin tones. This is also the case for the English language \parencite{1995ShannonSpeech}, where even envelopes modulating mostly noise were still capable of conveying meaning.

The envelope also helps to impart emotion and identity to the human voice \parencite{2018ZhuContributions}, and envelope preserving characteristics in concert halls are associated with their pleasantness \parencite{2011LokkiEngaging}.

@@

The problem of envelope detection, normally associated with the field of digital signal processing, can be made equivalent to the geometric problem of defining the (outer) shape of a set of points in $ \mathbb{R}^2 $, provided a method for addressing the difference of units in the horizontal, time related and vertical, amplitude related, axes of a digital signal is available, rendering the original DSP problem readily approachable via geometric techniques.

One direct way of defining the shape of a set of points in $ \mathbb{R}^2 $ is using the convex hull: as defined in \textcite{2008BergComputational}, the convex hull of a set of points $ S $ in two dimensional space is the smallest convex subset of the plane that contains all the points in $ S $. Convex hulls are unique, in the sense that a set of points define one and only one convex hull.

Generalizing that definition, \textcite{1983Edelsbrunnershape} introduced the concept of alpha-shapes, that can be seen as a Concave Hull; a mathematically well-defined generalization of the Convex Hull of a finite set of points, closely related to the Delaunay triangulation and Voronoi diagrams of those points.

The intuition behind the alpha shapes algorithm as that only points that can be touched by a circle of radius $ \alpha $ coming from an infinite distance towards the set of points of interest are part of the frontier, that is, of the Concave Hull of this set.

Once an $ \alpha $ parameter is defined alpha shapes conserve the uniqueness property, besides reducing to the Convex Hull when $ \alpha = \infty $. Consider, for example, the artificially generate discrete wave in figure \ref{fig:Hulls}
below, where the links between consecutive samples were deemphasised to highlight the "set of points" nature of a signal: The Convex Hull and a Concave Hull, in the form of an alpha shape with a particular value of $ \alpha $ are illustrated.

\begin{figure}[ht!]
  \centering
    \includegraphics{images/00ConvexHull.pdf}
  \caption{A discrete signal interpreted as a set of points in the Cartesian plane, and the Convex Hull and Concave Hull at $ \alpha=300 $ defined by it}
  \label{fig:Hulls}
\end{figure}

The alpha shapes approach is used in areas such as the detection of features in images \parencite{2016VarytimidisAlpha}, reconstruction of surfaces from a cloud of points \parencite{2015WuAutomated} and Spectroscopy \parencite{2019XuModeling}, with the last work, which involves the estimation and removal of the Blaze function (a kind of envelope) of an echelle spectrograph, being particularly illustrative of the potential of synergy between geometric and DSP approaches.

Other steps in the direction of translating geometric algorithms to the context of envelope detection were also made by \textcite{2015YangSkeleton} via an algorithm based on the construction of a skeleton underlying the digital wave of interest, and also via the direct translation of computer vision methods to the task of envelope detection \parencite{2015YangRepresenting}.

Following this path, we present a general approach to envelope detection, exploiting the intrinsic characteristics of a generic, spectrally complex wave, in order to avoid as much as possible the need for manual intervention or parameter tuning, such as the choice of $ \alpha $.

% <|<|<|<|<| TODO

@@

\section{Characterization of the temporal envelope}

The interest in the theory of envelope detection arose in the analogue domain, with the widespread adoption of radio communications, where the problem is typically restricted to simple sinusoids, often with well defined frequencies \parencite{2011TurnerDemodulation}.

In this context, the most mathematically sound definition of an envelope involves the representation of its underlying wave as an analytic signal, introduced by Gabor in 1946 \parencite{2007HahnHistory}. In his work, \textcite{1946GaborTheory} applies the then relatively new mathematical machinery of the quantum mechanics to unify the time and frequency domain representations of a wave, showing how the Hilbert transform could be applied to a real signal in order to obtain an equivalent complex signal, later known as the analytic signal.

Despite its widespread use, and effectiveness in the context of narrowband signals, envelope detection techniques based on the Hilbert transform don't behave well for broadband signals \parencite{2012DauSpeech}, being even physically
paradoxical in some cases \parencite{1996Loughlinamplitude}.

Figure \ref{fig:AnalyticSignal} exemplifies that, showing the envelope obtained via the analytic signal, for a pure sinusoid with a local frequency of 20 cycles, modulated by a polynomial of the third degree, illustrating the difference in the shape of the envelope in the presence and absence of Gaussian white noise with standard deviation of $ 1/10 $ of the wave's maximum amplitude.

It is readily noticeable from figure \ref{fig:AnalyticSignal} that, as soon as noise is introduced in the original signal, it is reflected in the envelope.

\begin{figure}[ht!]
  \centering
    \includegraphics[width=0.8\linewidth]{02AnalyticSignal.pdf}
  \caption{Envelope of a pure sinusoid modulated by a polynomial of degree 3, as obtained by the Hilbert Transform approach. In the first half, the sinusoid is free of noise, while in the second half white gaussian noise with a standard deviation of $ 1/10 $ of the maximum amplitude of the wave was added to the base signal.}
  \label{fig:AnalyticSignal}
\end{figure}

In general, the problem of envelope detection, or demodulation, can be interpreted as the task of, given a continuous wave $ W(t) $, decomposing it in two components $ E(t) $ and $ C(t) $, such that $ W(t) = E(t) C(t) $ \parencite{2011TurnerDemodulation}. 
$ E(t) $ represents the slow varying part of the wave, also known as the (temporal) envelope, modulator component, or amplitude modulation (AM) of $ W $, while $ C(t) $ models its fast varying part, called throughout the literature as its (temporal) fine structure, carrier component, or frequency modulation (FM).

In the case of broadband signals, the problem of envelope detection of an arbitrary wave is ill posed, in the sense that an infinite number of pairs of $ E(t) $ and $ C(t) $ can result in the same specific $ W(t) $ \parencite{2011TurnerDemodulation, 2013Mengempirical, 1996Loughlinamplitude}.

We address this problem by assuming $ C(t) $ normalized between $ \{-1, 1\}$. This assumption doesn't cause any loss of generality since, given an arbitrary $ C(t) $, we can obtain its normalization by dividing the function by its absolute global maximum, that is $ \hat{C}(t) = C(t) / \max(|C(t)|) $ provided that $ \max(|C(t)|) != 0 $.

Also, in this work we are concerned with the discrete version of this problem: given a finite digital wave represented by the vector $ W \in \mathbb{R}^n $, obtaining the temporal envelope of $ W $. The preceding definitions can be translated to a discrete setting assuming that the discrete quantities arise from observing the continuous ones at regular time intervals. In that case, the equality $ i = t \ \mathit{fps} $ can be used, where $ i $ is the index of each observation, $ t $ stands for the time in seconds, and $ \mathit{fps} $ is the frame rate, or the number of observations made in the period of a second.

% <|<|<|<|<| TODO We can thus define:

\section{Methodology}
% Overview of methods

 Since $ W $ is a wave, we expect it to be quasi-periodic, as well as uniformly sampled.

Instrumental to the method here presented is the definition of a pulse $ P \subset W $: each time $ W $ crosses the horizontal axis, the beginning of a pulse is defined, with the next crossing defining its end.

This definition is in line with the one presented in \textcite{national1996telecommunications}, where a pulse is defined as
a rapid change in the amplitude of a signal, followed by a fast return to the baseline value.

With a pulse thus defined, we can interpret a wave as composed of a succession of alternating positive and negative pulses.

As such, some structures can be defined, as illustrated in figure \ref{fig:Pulses}.

$ W = \{w_0, w_1, \cdots, w_{n-1}\} $

% <|<|<|<|<|

Half of a pseudo cycle, then, can be defined as the maximum set of adjacent samples with the same sign. Two adjacent half pseudo cycles form a full pseudo cycle, with the whole wave being formed by groups of adjacent pseudo cycles. The wave in figure \ref{fig:Pulses} if formed by 100 samples, that can be divided in 7 half pseudo cycles, of which two - the leftmost and the rightmost - are malformed, since they are not completely observable in the data provided by $ W $.

With the remaining half pseudo cycles, two full pseudo cycles can be formed. Given the quasi-periodicity of the digital wave $ W $, the order in which two adjacent half pseudo cycles are joined to form a full pseudo cycle can be chosen arbitrarily, provided consistency is maintained; in figure \ref{fig:Pulses}, a positive half pseudo cycle is followed by the adjacent negative one, forming a pseudo cycle.

\begin{figure}[ht!]
  \centering
    \includegraphics{images/01Pulses.pdf}
  \caption{Example of a discrete wave. The black circles are the individual samples of the signal, representing the intensity of the wave at regular time intervals. The grey areas are the half pseudo cycles. The dashed lines mark the divisions between adjacent pseudo cycles.}
  \label{fig:Pulses}
\end{figure}

From figure \ref{fig:Pulses} and keeping in mind the fact that the variation of the temporal envelope is much slower than the temporal fine structure variation, it is reasonable to suppose that the envelope would ideally pass through the point of maximum amplitude of each half pseudo cycle, following the least sinuous trajectory possible, while still tangencing each half pseudo cycle.

% <|<|<|<|<|

@@

The best representation one can hope to get of the original wave is, then, the digital wave itself. It is sensible, thus, to characterize the envelope of the wave in terms of a subset of the exact samples that comprise its digital representation.

%
